An Axiomatic Basis for Computer Programming: Review

### 2017-02-15 hoare/paper.tex
Continue of equation example, awesome conclusions

So Hoare suggested triple \{P\} Q \{R\} \cite{hoare:69} in which developer need
define 
precondition (premises) and postcondition (conclusion) which hold for program Q
after execution.

This technique helps prove correctness of a program Q, according to defined
logical predicates P and R.

It doesn't help derive true predicates from requirements, but it
provide ability to use common proof techniques for deducting correctness.

For next disscussion suppose that we define right P and R from some
requirements. \\

@@

Using axioms and inference rules what was proposed by Hoare, we could deduct
that if P holds before execution of Q, and if Q terminates, then R will holds
after execution (correctness of Q).\\

In 1969 paper Hoare didn't mentioned that Q should terminates, so he defined
rules just for deducting partial correctness (particular in rule D3). But later
he add definition for total correctness too. \\

@@

We limits input domain as set of real numbers and according to discriminant in
predicate P (suppose we used general quadratic formula):
$$ P \equiv (a, b, c \in \mathbb{R}) \wedge (b^2 - 4ac \geq 0)$$

Then we could find values by some program Q.
And finally check whether equation holds or not in R: 
$$ R \equiv (X \subseteq \mathbb{R}) \wedge (\forall x \in X \bullet ax^2 + bx + c
= 0)$$

@@

In the article Hoare states four basic rules:
\begin{itemize}
\item
  Axiom of Assignment
\item
  Rule of Consequence
\item
  Rule of Composition
\item
  Rule of Iteration
\end{itemize}

Also it requires no side effects for Q (nothing changes except state which
we define in pre and postconditions). Because in other case, we couldn't
consider correctness of overall state. \\

Using this rules we can make deductive proof from our premise to conclusion,
that R will holds after execution.

@@

It could be a case when program Q during execution
change state which doesn't hold conditions P and R.

@@

If we think about it in more details, we could decompose it to two sequence of
statements (rule of composition D2), in which first part hold some postcondition
R1 and second have R1 as precondition.

So nothing scary in that case if we still could prove correctness of our program.

P like a guard, filter input on what program will work correctly, and R check
results to be correct.

@@

However precondition and postconditions could be almost same. This looks ugly.

firstly we should define the
truth.

So directed tests could show that implementation isn't work properly. This is
helpful during development and evolving software product.

>>>>

So Hoare suggested triple $\{P\} Q \{R\}$ % \cite{hoare:69}
in which developer need
define precondition (premises) and postcondition (conclusion) which hold for
program $Q$ after execution.

This technique helps prove correctness of a program $Q$, according to defined
logical predicates $P$ and $R$.

It doesn't help derive true predicates from requirements, but it
provide ability to use common proof techniques for deducting correctness.
In other words, it helps us check that program do things right, but doesn't
affirm that what we want, is right things.

For next discussion suppose that we define right $P$ and $R$ from some
requirements. \\

@@

Using axioms and inference rules what was proposed by Hoare, we could deduct
that if $P$ holds before execution of $Q$, and if $Q$ terminates, then $R$ will
holds after execution (correctness of $Q$).

In 1969 paper Hoare didn't mentioned that $Q$ should terminates, so he defined
rules just for deducting partial correctness (particular in rule D3). But later
he add definition for total correctness too.

@@

We restrict input domain as set of real numbers and according to discriminant in
predicate $P$ (suppose we used general quadratic formula):
$$ P \equiv (a, b, c \in \mathbb{R}) \wedge (b^2 - 4ac \geq 0)$$

Then we could find values by some program $Q$. Suppose as a result program
returns set of values $X$ (Don't think about implementation yet).
And finally check whether equation holds or not in R: 
$$ R \equiv (X \subseteq \mathbb{R}) \wedge (\forall x \in X \bullet ax^2 + bx + c
= 0)$$

$P$ like a guard, filter input on what program will work correctly, and $R$
check results to be correct.

@@

In the '69 article Hoare states four basic rules:
\begin{itemize}
\item
  Axiom of Assignment
\item
  Rule of Consequence
\item
  Rule of Composition
\item
  Rule of Iteration
\end{itemize}
But sometimes another two rules is stated - Skip Axiom and Condition Rule. First
one could be derived from Axiom of Assignment, and second from Rule of
Iteration.\\

All this rules implicitly assumed the absence of side effects for $Q$ (nothing
changes except state which we define in pre and postconditions). Because in
other case, we couldn't consider correctness of overall state. \\

Using this rules we can make deductive proof from our premise to conclusion,
that $R$ will holds after execution.

@@

It could be a case when program $Q$ during execution
change state which doesn't hold conditions $P$ and $R$.

@@

If we think about it in more details, we could decompose it to two sequence of
statements (by rule of composition D2), in which first part hold some postcondition
$R_1$ and second have $R_1$ as precondition. But $R_1$ could be arbitrary, for example
$\neg P$.

As we can see, there is nothing scary in this case if we still could prove
correctness of our program. \\

Lets return to our example about real roots.

Suppose we choose general quadratic formula for solving equation:
$$ x_{1,2} = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}$$

We know that $b^2 - 4ac$ not negative, but we
don't know weather $a$ is equals to zero.
In this case equation became linear, but still have solutions.

Suppose we didn't see that general formula is derived from equation. Lets just
apply substitution all occurrence of $x$ in postcondition by stated expressions.

After polynomial expansion we could see that equation takes the form:
$$\frac{b^2}{2a} \mp \frac{b\sqrt{b^2 - 4ac}}{2a} - c - \frac{b^2}{2a} \pm
\frac{b\sqrt{b^2 - 4ac}}{2a} + c = 0$$

I little bit cheated, because I already reduce some parts by $a$, but
nevertheless, after reducing the remaining we get:
$$ 0 = 0 $$

But we should remember, that we could reduce only by not zero value, thus
postcondition holds only if $a \neq 0$.

Lets predicate $B \equiv a \neq 0$, then precondition for our statement:
$$ P_1 \equiv P \wedge B $$

By the manipulations above we actually proved that our statement correct, with
respect to precondition $P_1$ and postcondition $R$.

Actually we could replace whole program with one assignment:
$$ X = \emptyset $$

And it will be correct program, according to specification. That's why I
stressed that it doesn't help obtain right predicates.

@@

% However precondition and postconditions could be almost same. This looks ugly.

% firstly we should define the
% truth.

% So directed tests could show that implementation isn't work properly. This is
% helpful during development and evolving software product.
